---
title: "Lab 5"
format: html
editor: visual
embed-resources: true
---

## Lab 5

**Setup**

```{r}
# Where are we getting the data from
met_url <- "https://github.com/USCbiostats/data-science-data/raw/master/02_met/met_all.gz"

# Downloading the data to a tempfile (so it is destroyed afterwards)
# you can replace this with, for example, your own data:
tmp <- tempfile(pattern = "met", fileext = ".gz")
# tmp <- "met.gz"

# We should be downloading this, ONLY IF this was not downloaded already.
# otherwise is just a waste of time.
if (!file.exists(tmp)) {
  download.file(
    url      = met_url,
    destfile = tmp,
    # method   = "libcurl", timeout = 1000 (you may need this option)
  )
}

# Reading the data
met <- read.csv(tmp)
head(met)


```

```{r}
# Download the data
stations <- read.csv("https://noaa-isd-pds.s3.amazonaws.com/isd-history.csv")
stations$USAF <- as.integer(stations$USAF)

# Dealing with blanks and 999999
stations$USAF[stations$USAF == 999999] <- NA
stations$CTRY[stations$CTRY == ""] <- NA
stations$STATE[stations$STATE == ""] <- NA

# Selecting the three relevant columns, and keeping unique records
stations <- unique(stations[, c('USAF', 'CTRY', 'STATE')])

# Dropping NAs
stations <- stations[!is.na(stations$USAF), ]

# Removing duplicates
stations <- stations[!duplicated(stations$USAF), ]

# Merging Data 
met <- merge(
  met,
  stations,
  by.x = "USAFID",
  by.y = "USAF",
  all.x = TRUE
)

head(met)
```

**Question 1.**

```{r}
library(dplyr)

# Calculate the average values for each continental US station
station_summary <- met %>%
  filter(
    CTRY == "US",
    !STATE %in% c("AK", "HI"),
    !is.na(STATE)
  ) %>%
  group_by(USAFID) %>%
  summarise(
    STATE = first(STATE),
    avg_temp = mean(temp, na.rm = TRUE),
    avg_wind = mean(wind.sp, na.rm = TRUE),
    avg_pressure = mean(atm.press, na.rm = TRUE),
    .groups = "drop"
  )

# Find the median of the station averages
median_temp <- quantile(station_summary$avg_temp, 0.5, na.rm = TRUE)
median_wind <- quantile(station_summary$avg_wind, 0.5, na.rm = TRUE)
median_pressure <- quantile(station_summary$avg_pressure, 0.5, na.rm = TRUE)

# Find the station closest to each median
temp_station <- station_summary %>%
  filter(is.finite(avg_temp)) %>%
  slice_min(abs(avg_temp - median_temp), n = 1, with_ties = FALSE)

wind_station <- station_summary %>%
  filter(is.finite(avg_wind)) %>%
  slice_min(abs(avg_wind - median_wind), n = 1, with_ties = FALSE)

pressure_station <- station_summary %>%
  filter(is.finite(avg_pressure)) %>%
  slice_min(abs(avg_pressure - median_pressure), n = 1, with_ties = FALSE)

# Display the answers
temp_station
wind_station
pressure_station
```

The station closest to the median temperature was station 720458 in Kentucky, with an average temperature of 23.7 Celsius. The station closest to the median wind speed was 720929 in Wisconsin, with an average wind speed of 17.4 m/s. The station closest to the median pressure was station 722238 in Alabama, with an average pressure of 1015 hPa. All three stations were different.

**Question 2**

```{r}
# Compute median latitude and longitude for each state
state_medians <- met %>%
  filter(
    CTRY == "US",
    !STATE %in% c("AK", "HI"),
    !is.na(STATE)
  ) %>%
  distinct(USAFID, STATE, lat, lon) %>%
  group_by(STATE) %>%
  summarize(
    med_lat = median(lat, na.rm = TRUE),
    med_lon = median(lon, na.rm = TRUE)
  )

state_medians
```

**Question 3**

```{r}
# Keep one row per station
stations_unique <- met %>%
  filter(
    CTRY == "US",
    !STATE %in% c("AK", "HI"),
    !is.na(STATE),
    !is.na(lat),
    !is.na(lon)
  ) %>%
  distinct(USAFID, STATE, lat, lon)

# Find the station closest to each state's mean coordinates
mean_stations <- stations_unique %>%
  group_by(STATE) %>%
  mutate(
    mean_lat = mean(lat),
    mean_lon = mean(lon),
    distance = sqrt(
      (lat - mean_lat)^2 + (lon - mean_lon)^2
    )
  ) %>%
  slice_min(distance, n = 1, with_ties = FALSE) %>%
  ungroup()

mean_stations

library(leaflet)

leaflet() %>%
  addTiles() %>%
  
  # Median coordinates from Question 2
  addCircleMarkers(
    data = state_medians,
    lng = ~med_lon,
    lat = ~med_lat,
    color = "blue",
    radius = 5,
    label = ~paste(STATE, "- median coordinates")
  ) %>%
  
  # Stations closest to the mean coordinates
  addCircleMarkers(
    data = mean_stations,
    lng = ~lon,
    lat = ~lat,
    color = "red",
    radius = 5,
    label = ~paste(STATE, "- station", USAFID)
  ) %>%
  
  addLegend(
    position = "bottomright",
    colors = c("blue", "red"),
    labels = c(
      "Median coordinates",
      "Station closest to mean coordinates"
    )
  )
```

**Question 4**

```{r}
met_categories <- met %>%
  filter(
    CTRY == "US",
    !STATE %in% c("AK", "HI"),
    !is.na(STATE)
  ) %>%
  group_by(STATE) %>%
  mutate(
    state_avg_temp = mean(temp, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  mutate(
    temp_category = case_when(
      state_avg_temp < 20 ~ "Low",
      state_avg_temp >= 20 & state_avg_temp < 25 ~ "Mid",
      state_avg_temp >= 25 ~ "High"
    )
  )

state_summary <- met_categories %>%
  group_by(temp_category) %>%
  summarise(
    number_states = n_distinct(STATE),
    number_entries = n(),
    number_NA_entries = sum(is.na(temp)),
    number_stations = n_distinct(USAFID),
    mean_temperature = mean(temp, na.rm = TRUE),
    mean_wind_speed = mean(wind.sp, na.rm = TRUE),
    mean_pressure = mean(atm.press, na.rm = TRUE)
  )

state_summary
```
